On the Limitations of Weighted Model Counting

### 2020-03-11 paper.tex
Some new results on extending BAs to make measures WMC-computable

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{bm}
\usepackage{booktabs}
\usepackage[backgroundcolor=lightgray]{todonotes}

@@

\title{Weighted Model Counting as a Special Case of Polyadic Measure Algebras}
\author{Paulius Dilkas}

@@

\section{Propositional Logic and Boolean Algebras}

@@

\section{First-Order Logic and Polyadic Algebras}

@@

\section{From First-Order Logic to Polyadic Algebras}

@@

% TODO: maybe I should gives this kind of a BA a name? A synonym of 'complete',
% perhaps.
% TODO: a special case for weight=0.
\begin{proposition}
  Let $\mathbf{B}$ be a finite measure algebra with measure $m : \mathbf{B} \to
  \mathbb{R}_{\ge 0}$. Let $L \subset \mathbf{B}$ be defined as
  \[
  L = \{ l_i \mid i \in [n] \} \cup \{ \neg l_i \mid i \in [n] \}
  \]
  for some $n \in \mathbb{N}$. Finally, assume that $\mathbf{B}$ has $2^n$
  atoms, where each atom $a \in \mathbf{B}$ is an infimum
  \[
    a = \bigwedge_{i=1}^n a_i
  \]
  such that $a_i \in \{ l_i, \neg l_i \}$ for $i \in [n]$. Then there exists a
  weight function $w : L \to \mathbb{R}_{>0}$ that makes $m$ a WMC measure if
  and only if
  \begin{equation} \label{eq:wmccondition}
  m(l \land l') = m(l)m(l')
  \end{equation}
  for all distinct $l, l' \in L$ such that $l \ne \neg l'$.
\end{proposition}

@@

  Now recall that $m = \nWMC$ and note that \cref{eq:to_prove2} can be
  multiplied by $\WMC(1)^2$ to turn the equation into one for $\WMC$ instead of
  $\nWMC$. Then
  \begin{align*}
    \WMC(k_i \land k_j) &= \sum_{\text{atoms } a \le k_i \land k_j} \WMC(a) = \sum_{\text{atoms } a \le k_i \land k_j} \prod_{m \in [n]} w(a_m) \\
                        &= \sum_{\text{atoms } a \le k_i \land k_j} w(a_i)w(a_j) \prod_{m \in [n] \setminus \{ i, j \}} w(a_m) = \sum_{\text{atoms } a \le k_i \land k_j} w(k_i)w(k_j) \prod_{m \in [n] \setminus \{ i, j \}} w(a_m) \\
    &= w(k_i)w(k_j) \sum_{\text{atoms } a \le k_i \land k_j} \prod_{m \in [n] \setminus \{ i, j \}} w(a_m) = w(k_i)w(k_j)C,
  \end{align*}
  where $C$ denotes the part of $\WMC(k_i \land k_j)$ that will be the same for
  $\WMC(\neg k_i \land k_j)$, $\WMC(k_i \land \neg k_j)$, and $\WMC(\neg k_i
  \land \neg k_j)$ as well. But then \cref{eq:to_prove2} becomes
  \[
    w(k_i)w(k_j)w(\neg k_i)w(\neg k_j)C^2 = w(k_i)w(\neg k_j)w(\neg k_i)w(k_j)C^2
  \]
  which is trivially true. By showing that WMC satisfies \cref{eq:to_prove2}, we
  also showed that it satisfies \cref{eq:to_prove}, finishing the second part of
  the proof.
  \todo[inline]{Adapt the proof to zero measure events}
\end{proof}

\todo[inline,caption={}]{The big TODO list
  \begin{itemize}
  \item Extension to infinite (atomic?) Boolean algebras.
  \item How many extra variable do you need to add to make any probability
    distribution representable using WMC?
  \item Abstraction refinements as homomorphisms.
  \item Definition of a measure-preserving homomorphism from Jech's set theory
    book.
  \item A Boolean algebra is approximable if its Stone space is approximable.
  \end{itemize}
}

\section{Homomorphisms}

@@

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage[backgroundcolor=lightgray]{todonotes}

@@

\usetikzlibrary{cd}

\tikzset{
  Subset/.style={
    draw=none,
    every to/.append style={
      edge node={node [sloped, allow upside down, auto=false]{$\subset$}}}
  }
}

\title{On the Limitations of Weighted Model Counting}
\author{Paulius Dilkas}

@@

\section{WMC as a Measure}

@@

\section{An Extension to First-Order Logic}

@@

\section{How Probabilities Are Computed}

@@

\section{What Measures Are WMC-Computable?}

\subsection{WMC Requires Independent Literals}

% TODO: maybe I should gives this kind of a BA a name? A synonym of 'complete',
% perhaps.
% TODO: a special case for weight=0.

\begin{proposition}
  Let $\mathbf{B}$ be a finite measure algebra with measure $m : \mathbf{B} \to
  \mathbb{R}_{\ge 0}$. Let $L \subset \mathbf{B}$ be defined as
  \[
  L = \{ l_i \mid i \in [n] \} \cup \{ \neg l_i \mid i \in [n] \}
  \]
  for some $n \in \mathbb{N}$. Finally, assume that $\mathbf{B}$ has $2^n$
  atoms, where each atom $a \in \mathbf{B}$ is an infimum
  \[
    a = \bigwedge_{i=1}^n a_i
  \]
  such that $a_i \in \{ l_i, \neg l_i \}$ for $i \in [n]$. Then there exists a
  weight function $w : L \to \mathbb{R}_{>0}$ that makes $m$ a WMC measure if
  and only if
  \begin{equation} \label{eq:wmccondition}
  m(l \land l') = m(l)m(l')
  \end{equation}
  for all distinct $l, l' \in L$ such that $l \ne \neg l'$.
\end{proposition}

@@

  Now recall that $m = \nWMC$ and note that \cref{eq:to_prove2} can be
  multiplied by $\WMC(1)^2$ to turn the equation into one for $\WMC$ instead of
  $\nWMC$. Then
  \begin{align*}
    \WMC(k_i \land k_j) &= \sum_{\text{atoms } a \le k_i \land k_j} \WMC(a) = \sum_{\text{atoms } a \le k_i \land k_j} \prod_{m \in [n]} w(a_m) \\
                        &= \sum_{\text{atoms } a \le k_i \land k_j} w(a_i)w(a_j) \prod_{m \in [n] \setminus \{ i, j \}} w(a_m) = \sum_{\text{atoms } a \le k_i \land k_j} w(k_i)w(k_j) \prod_{m \in [n] \setminus \{ i, j \}} w(a_m) \\
    &= w(k_i)w(k_j) \sum_{\text{atoms } a \le k_i \land k_j} \prod_{m \in [n] \setminus \{ i, j \}} w(a_m) = w(k_i)w(k_j)C,
  \end{align*}
  where $C$ denotes the part of $\WMC(k_i \land k_j)$ that will be the same for
  $\WMC(\neg k_i \land k_j)$, $\WMC(k_i \land \neg k_j)$, and $\WMC(\neg k_i
  \land \neg k_j)$ as well. But then \cref{eq:to_prove2} becomes
  \[
    w(k_i)w(k_j)w(\neg k_i)w(\neg k_j)C^2 = w(k_i)w(\neg k_j)w(\neg k_i)w(k_j)C^2
  \]
  which is trivially true. By showing that WMC satisfies \cref{eq:to_prove2}, we
  also showed that it satisfies \cref{eq:to_prove}, finishing the second part of
  the proof.
\end{proof}
% TODO: the auxiliary result should be a 'claim' in-between the theorem and the proof.

\subsection{Extending the Algebra}

\subsubsection{Preliminaries}

@@

\subsubsection{New Results}

A well-known way to overcome this limitation of independence is by adding more
literals \cite{DBLP:journals/ai/ChaviraD08}, i.e., extending the set $L$ covered
by the WMC weight function $w : L \to \mathbb{R}_{>0}$. Let us translate this
idea to the language of Boolean algebras.

\begin{theorem} \label{thm:extension} % TODO: cite the fact about atoms
  Let $\mathbf{B}$ be a finite Boolean algebra freely generated by some set of
  `literals' $L$, and let $m : \mathbf{B} \to \mathbb{R}_{\ge 0}$ be an
  arbitrary measure. We know that $\mathbf{B}$ has $n = 2^{|L|}$ atoms. Let
  $(a_i)_{i=1}^n$ be those atoms. Let $L' = L \cup \{ \phi_i \mid i \in [n] \}$
  be the set $L$ extended with $n$ new literals. Let $\mathbf{B'} =
  \mathbf{C}/I$ be the quotient of the Boolean algebra $\mathbf{C}$ freely
  generated by $L'$ and the ideal $I$ generated by
  \[
    \{\phi_i \land \phi_j \mid i, j \in [n], i \ne j \} \cup \{ a_i \land \phi_j
    \mid i, j \in [n], i \ne j \}.
  \]
  Let $h : \mathbf{B} \to \mathbf{B'}$ be the Boolean homomorphism defined by
  $h(l) = l$ for all $l \in L$ (note that this indeed defines a homomorphism).
  Let $w : L' \to \mathbb{R}_{> 0}$ be defined by
  \[
    w(l) = \begin{cases}
      m(a_i) & \text{if } l = \phi_i \text{ for some } i \in [n] \\
      1 & \text{otherwise}
    \end{cases}
  \]
  for all $l \in L'$. This defines a WMC measure $m' : \mathbf{B'} \to
  \mathbb{R}_{\ge 0}$. Furthermore,
  \[
    m(a) = m'(h(a)).
  \]
\end{theorem}

In simpler terms, any measure can be computed using WMC by extending the Boolean
algebra with more literals and connecting the two via a homomorphism. More
precisely, we are given the red part in
\[
  \begin{tikzcd}
    & \textcolor{red}{\mathbb{R}_{\ge 0}} & & \\
    \textcolor{red}{\mathbf{B}} \arrow[red]{ur}{m} \arrow{rr}{h} & & \mathbf{B'}
    \arrow{ul}[swap]{m'} & \\
    \textcolor{red}{L} \arrow[Subset,red]{u}{} \arrow[Subset]{rr}{} & & L'
    \arrow[Subset]{u}{} \arrow{r}{w} & \mathbb{R}_{>0}
  \end{tikzcd}
\]
and construct the black part in such a way that the triangle commutes.

\begin{proof}
  % TODO
\end{proof}

% TODO: proof that n is the smallest number (in the worst case, without
% independence assumptions)
%\begin{proposition}
%  In \cref{thm:extension}, if $L'$ has less than $n$ additional elements, then
%  there exists a measure $$
%\end{proposition}

% TODO: describe how my approach is better in the worst case than that by
% Darwiche

\todo[inline,caption={}]{The big TODO list
  \begin{itemize}
  \item Extension to infinite (atomic?) Boolean algebras.
  \item Compare my polyadic measures with first-order WMC.
  \item Abstraction refinements as homomorphisms.
  \item Definition of a measure-preserving homomorphism from Jech's set theory
    book.
  \item A Boolean algebra is approximable if its Stone space is approximable.
  \end{itemize}
}
